Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021

### 2024-01-25 outputs/paper/paper.qmd
updated charts

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism to counter the injustices related to police power abuse around strip searching of the vulnerable Toronto population. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using stacked summary tables, stacked bar charts and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching for intersecting social identities With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

This chosen dataset was published by Toronto Police Services to convey information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features information on date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with details on any items found. Informative as it is, the dataset contains much naming inconsistency and many columns outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

To streamline the data regarding racial and gender identities of the people involved in police strip searches, the names within the original dataset are simplified and then the dataset is split into a smaller one containing relevant columns ([see @tbl-cleaned_race_gender_data]). This dataset contains demographic and strip search-relevant information for each police arrest event, such as race, gender, whether the event involves a strip search, and if it is a strip search, are there any items found.

@@

  cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = race, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm'), axis.text.x = element_text(angle = 70, vjust = 1, hjust= 1)) +
  labs(x = "Race",
       y = "Number of Arrests",
       fill = "Strip Search"
  ) +
  scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  filter(strip_search == "Yes") |> 
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar()+theme_minimal()+ylab("number of arrests")+facet_wrap(~race)+ scale_fill_viridis_d(option="cividis")

@@

  cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = gender, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Gender",
       y = "Number of Arrests",
       fill = "Strip Search"
  ) +
  scale_fill_viridis_d(option="cividis")

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-reasons-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

@@

#create a scatterplot to show each observation in the dataset, comparing the total reasons to racial identitiy found in each strip search
#colour the points based on the perceived race of the captive
#code for geom_point() referenced from https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto/blob/main/Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd
cleaned_search_reasons_data |>
  group_by(total_reasons, race) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = race, size = count)) +
  geom_point(alpha = 0.5) +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Race"
       ) +
  scale_fill_viridis_d(option="cividis")

@@

cleaned_search_reasons_data |>
  group_by(total_reasons, gender) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = gender, size = count)) +
  geom_point(alpha = 0.5) +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Gender"
       ) +
  scale_fill_viridis_d(option="cividis")

```
## Discussion

>>>>

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism, contributing to the dialogue around police power abuse, using strip-search to undermine the the vulnerable populations in Toronto. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using stacked summary tables, stacked bar charts and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching for intersecting social identities With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

This chosen dataset was published by Toronto Police Services to convey information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features information on date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with any items found. Data of strip searches are lodged based on each police arrest throughout 2020 and 2021, using a unique identifier called `event_id` which identifies the nature of the strip search and the social identities of the people involved.

Informative as it is, the dataset contains much naming inconsistency and many columns outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

Most relevant to the paper are the variables on the racial and gender identities of people who were arrested and then strip-searched. To streamline this data, the names within the original dataset are simplified and then the dataset is split into a smaller dataset containing relevant columns ([see @tbl-cleaned_race_gender_data]). This dataset entails demographic and strip search-relevant information for each police arrest event, such as race, gender, whether the event involves a strip search, and if it is a strip search, are there any items found.

@@

  cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = race, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") +
  theme(legend.key.size = unit(0.5, 'cm'), axis.text.x = element_text(angle = 70, vjust = 1, hjust= 1)) +
  labs(x = "Race",
       y = "Number of Arrests",
       fill = "Strip Search") + scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  filter(strip_search == "Yes") |> 
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar() + 
  labs(y = "Number of Arrests", x = "Strip Search", fill = "Strip Search") +
  facet_wrap(~race) + scale_fill_viridis_d(option="cividis")

@@

  cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = gender, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Gender",
       y = "Number of Arrests",
       fill = "Strip Search"
  ) + scale_fill_viridis_d(option="cividis")

@@

```{r}
#| echo: false
#| message: false
#| results: hide
#| label: tbl-reasons-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

@@

#create a scatterplot to show each observation in the dataset, comparing the total reasons to racial identitiy found in each strip search
#colour the points based on the perceived race of the captive
#code for geom_point() referenced from https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto/blob/main/Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd
cleaned_search_reasons_data |>
  group_by(total_reasons, race) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = race)) +
  geom_line(alpha = 0.5) +
  geom_point(aes(size = count), alpha = 0.5) +
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Race"
       ) +
  theme(legend.key.size = unit(0.5, 'cm')) +
  scale_fill_viridis_d(option="cividis")

@@

cleaned_search_reasons_data |>
  group_by(total_reasons, gender) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = gender)) +
  geom_line(alpha = 0.5) +
  geom_point(aes(size = count), alpha = 0.5) +
  labs(y = "Count", x = "Total Reasons for Strip Searching", color = "Gender") +
  theme(legend.key.size = unit(0.5, 'cm')) +
  scale_fill_viridis_d(option="cividis")

## Discussion
